Modal Abstractions for Virtualizing Memory Addresses

### 2025-07-18 background_C.tex

\end{tikzpicture}
          \caption{Accessing to the Page Referenced by L1 Entry}
        \label{fig:enter-label}
    \end{subfigure}
    \vfill
\begin{minipage}{.5\textwidth}
%    \begin{figure}
\begin{lstlisting}[mathescape,escapeinside={(*}{*)},basicstyle=\footnotesize]
/*
@param entry: entry whose frame address is used 
              to locate the subsequent entry in the page-tree
@return next: return a virtual pointer to the descendant tree
*/
pte_t *pte_get_next_table(pte_t *entry) {
  pte_t *next;
  /*Reading page table entry:
    pointed by a virtual memory address*/
  /* Jump if the present bit is not zero:
     no need to allocate next level (* \label{line:check_entry_presentC} *)(* \label{line:mask_presentC} *)  (*\label{line:after_concluding_qfrac1C}*)*/
  if (!entry->present(*\label{line:read_entry_contentsC}*) (*\label{line:conditional_childrenC}*)) {
    /*Allocation path starts here (*\label{line:pass_addrof_nextC}*)(*\label{line:alloc_path_startC}*)*/
    pte_initialize(entry);(* \label{line:call_to_pte_initializeC} *)
    /*Entry value updates:*/
    entry->pfn = nextpaddr;
    entry->present = 1; (*\label{line:install_new_entryC}*)
    /*Entry is initialized,
      so we satisfy the condition to access children list */
  } (*\label{line:end_of_allocation_pathCB}*)
  /*Physical to virtual address mapping code segment starts here (*\label{line:finalpieceSB}*)*/
  uintptr_t next_phys_addr = PTE_PFN_TO_ADDR(entry->pfn)
  uintptr_t next_virt_addr = (uintptr_t) P2V(next_phys_addr); (*\label{line:p2vCB}*);
  next = (pte_t *) next_virt_addr;(*\label{line:finalpieceEB}*)
  return next;
}
\end{lstlisting}
\end{minipage}
\begin{minipage}{.4\textwidth}
\begin{lstlisting}[mathescape,escapeinside={(*}{*)}, basicstyle=\footnotesize]
pte_t *walkpgdir(pte_t *root, void *va){ 
  pte_t *l4_entry = &root[L4Offset(va)];
  pte_t *l3 = pte_next_table(l4_entry);
  pte_t *l3_entry = &l3[L3Offset(va)];
  pte_t *l2 = pte_next_table(l3_entry);
  pte_t *l2_entry = &l2[L2Offset(va)];  
  pte_t *l1 = pte_next_table(l2_entry);
  pte_t *l1_entry = &l1[L1Offset(va)];  
  return l1_entry
}
\end{lstlisting}
%          \caption{Walking page-table tree to locate L1 entry C-source code.}
%        \label{fig:enter-label}
 %   \end{figure}
\end{minipage}
%\end{framed}
    \vspace{-2.5em}
    \caption{x86-64 page table lookups.}
    \label{fig:pagetables}
    \vspace{-1em}
\end{figure}

>>>>

\end{tikzpicture}
          \caption{Accessing to the Page Referenced by L1 Entry}
        \label{fig:enter-label}
    \end{subfigure}
    \vfill
\begin{minipage}{.5\textwidth}
%    \begin{figure}
\begin{lstlisting}[mathescape,escapeinside={(*}{*)},basicstyle=\footnotesize]
pte_t *pte_get_next_table(pte_t *entry) {
  pte_t *next;
  /*Reading page table entry:
    pointed by a virtual memory address*/
  /* Jump if the present bit is not zero:
     no need to allocate next level (* \label{line:check_entry_presentC} *)(* \label{line:mask_presentC} *)  (*\label{line:after_concluding_qfrac1C}*)*/
  if (!entry->present(*\label{line:read_entry_contentsC}*) (*\label{line:conditional_childrenC}*)) {
    /*Allocation path starts here (*\label{line:pass_addrof_nextC}*)(*\label{line:alloc_path_startC}*)*/
    pte_initialize(entry);(* \label{line:call_to_pte_initializeC} *)
    /*Entry value updates:*/
    entry->pfn = nextpaddr;
    entry->present = 1; (*\label{line:install_new_entryC}*)
    /*Entry is initialized,
      so we satisfy the condition to access children list */
  } (*\label{line:end_of_allocation_pathCB}*)
  /*Physical to virtual address mapping code segment starts here (*\label{line:finalpieceSB}*)*/
  uintptr_t next_phys_addr = PTE_PFN_TO_ADDR(entry->pfn)
  uintptr_t next_virt_addr = (uintptr_t) P2V(next_phys_addr); (*\label{line:p2vCB}*);
  next = (pte_t *) next_virt_addr;(*\label{line:finalpieceEB}*)
  return next;
}
\end{lstlisting}
\end{minipage}
\begin{minipage}{.4\textwidth}
\begin{lstlisting}[mathescape,escapeinside={(*}{*)}, basicstyle=\footnotesize]
pte_t *walkpgdir(pte_t *root, void *va){ 
  pte_t *l4_entry = &root[L4Offset(va)];
  pte_t *l3 = pte_next_table(l4_entry);
  pte_t *l3_entry = &l3[L3Offset(va)];
  pte_t *l2 = pte_next_table(l3_entry);
  pte_t *l2_entry = &l2[L2Offset(va)];  
  pte_t *l1 = pte_next_table(l2_entry);
  pte_t *l1_entry = &l1[L1Offset(va)];  
  return l1_entry
}
\end{lstlisting}
%          \caption{Walking page-table tree to locate L1 entry C-source code.}
%        \label{fig:enter-label}
 %   \end{figure}
\end{minipage}
%\end{framed}
    \vspace{-2.5em}
    \caption{x86-64 page table lookups.}
    \label{fig:pagetables}
    \vspace{-1em}
\end{figure}

### 2025-07-18 examples_C.tex

\subsubsection{Physical-to-Virtual Conversion with \textsf{P2V}}
\label{sec:p2vC}
Once we know the entry refers to a physical address in the identity mapping range ($\Xi$)
(via the branch at Line \ref{line:check_entry_present_jumpC}, or  by allocating and installing a new entry
as just discussed for Lines \ref{line:check_entry_present_jumpC}--\ref{line:end_of_allocation_pathC}), 
we can convert this frame address to a corresponding virtual address via the identity mappings
discussed in Section \ref{subsec:identitymappingsC} and Figure \ref{fig:peraspaceinvariant_with_p2v_extensionC}.
in the last lines of \lstinline|pte_get_next_table| shown in Figure \ref{fig:p2vC} (the continuation of Figure \ref{fig:calltopteinitializeC}).
This is a critical piece of the full page table walk verification.
In our small kernel (Line \ref{line:p2vC} in Figure \ref{fig:p2vC}), as in larger kernels, the C macro \texttt{P2V} common to many kernels
is actually just addition by the constant offset mentioned in Section \ref{subsec:identitymappingsC}.
But the correctness of this simple instruction is quite subtle.
%  and cannot be proven 
% without the extended invariant (Figure \ref{fig:peraspaceinvariant_with_p2v_extension})
% worked out Section \ref{subsec:identitymappings}.

%\begin{figure}\footnotesize
%  \begin{lstlisting}[mathescape,escapeinside={(*}{*)}]
%  /*... Continued from Figure (*\ref{fig:calltopteinitialize}*); assertions below specialized to non-allocating path for clarity */
%  $\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus \{ \mathsf{entry} \}),m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{rcx}  \mapsto_{\textsf{r}} \textsf{\_}  \ast \textsf{entry} \mapsto_{\textsf{id}} \textsf{\_} \ast \ghostmaptoken{\delta{}s}{\rtv}{\delta}  }_{\rtv}$
%  $\specline{ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{(pte\_initialized (entry\_val.pfn))} \urcorner }_{\rtv}$
%  $\specline{ \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{(pte\_initialized (entry\_val.pfn)))} \ast \texttt{rax} \mapsto_{\textsf{r}} \textsf{ table\_root (pte\_initialize(entry\_val.pfn))} }_{\rtv}$
%  $\specline{\forall_{i\in \textsf{0 ... 511} } \ldotp  \ghostmaptoken{\textsf{id}}{((\textsf{table\_root (pte\_initialized (entry\_val.pfn)))}) + \textsf{i * 8})}{\textsf{v-1}}  }$ (*\label{line:children}*)
%  uintptr_t next_phys_addr = PTE_PFN_TO_ADDR(entry->pfn)
%  uintptr_t next_virt_addr = (uintptr_t) P2V(next_phys_addr); (*\label{line:p2v}*) 
%  /*... clean up the stack and return the rax value*/
%  return next
%}
%\end{lstlisting}
%\vspace{-1em}
%\caption{Converting a physical address of a PTE to a virtual address (w/o instruction pointer or flag updates).
% % Abreviating the other relevant resources (including the lower entries) peeled-out of the invariant as $\textsf{R}_{\textsf{children}}$
%\vspace{-1em}
%}
%\label{fig:p2v}
%\end{figure}
Lines \ref{childrenC}-\ref{line:p2vC} in Figure \ref{fig:calltopteinitializeC} show the verification of the end of \lstinline|pte_get_next_table| specialized to the case where 
where no allocation was necessary (i.e., the conditional on Line \ref{line:check_entry_presentC} of Figure \ref{fig:calltopteinitializeC} was taken).
In this case, the true present bit allows access to the child tokens from Line \ref{line:conditional_childrenC} of Figure \ref{fig:calltopteinitializeC},
which is then refined to the assertion on Line \ref{line:childrenC} of Figure \ref{fig:p2vC}.
The code loads \lstinline|rcx| with the offset value \textsf{KERNBASE}, which gives us the value of the virtual address ($\textsf{entry}_{\textsf{pfn}}$ \textsf{+KERNBASE})
of the \emph{base} of the next level of the page table.
% \todo[inline]{the next sentence depends on having figure 10 updated to reflect the page-worth of tokens}
While we could now convert this address to a virtual points-to, this is not necessarily the correct thing to do.
The caller \lstinline|walkpgdir| (discussed next) uses \lstinline|pte_get_next_table| to retrieve just the base address,
because only the caller knows which entry in the subsequent table will be accessed (it depends on the corresponding bits from the virtual
address being translated). So instead we pass back the per-address-space invariant with the identity mapping resources for \lstinline|entry|
pulled out. The caller determines which entry in that table must actually
be accessed --- by selecting the appropriate index into the 512 ghost map tokens returned in the postcondition,
and using the ghost translation and physical location portions of the invariant to assemble a vpte-pointsto
that justifies the caller's subsequent access to a particular entry of the returned table.
% in the identity map ($\Xi\setminus\{entry\}$) of the kernel invariant.
% the logical update in Specification  Lines 5-10 to 10-14 for obtaining virtual-pointsto resource for the frame 
% ($\textsf{entry}_{\textsf{pfn}}$) by removing it from the ghost map ($\Xi\setminus\{entry\}\cup \{\textsf{entry}_{\textsf{pfn}}) \}$) 
% in Line 5 and compute the identity mapping for this physical frame address in Line 13 in Figure \ref{fig:p2v}).

>>>>

\subsubsection{Physical-to-Virtual Conversion with \textsf{P2V}}
\label{sec:p2vC}
Once we know the entry refers to a physical address in the identity mapping range ($\Xi$)
(via the branch at Line \ref{line:check_entry_present_jumpC}, or  by allocating and installing a new entry
as just discussed for Lines \ref{line:check_entry_present_jumpC}--\ref{line:end_of_allocation_pathC}), 
we can convert this frame address to a corresponding virtual address via the identity mappings
discussed in Section \ref{subsec:identitymappingsC} and Figure \ref{fig:peraspaceinvariant_with_p2v_extensionC}.
in the last lines of \lstinline|pte_get_next_table| shown in Line \ref{line:p2vC} in Figure \ref{fig:calltopteinitializeC} (the continuation of Figure \ref{fig:calltopteinitializeC}).
This is a critical piece of the full page table walk verification.
In our small kernel (Line \ref{line:p2vC} in Figure \ref{fig:calltopteinitializeC}), as in larger kernels, the C macro \texttt{P2V} common to many kernels
is actually just addition by the constant offset mentioned in Section \ref{subsec:identitymappingsC}.
But the correctness of this simple instruction is quite subtle.
%  and cannot be proven 
% without the extended invariant (Figure \ref{fig:peraspaceinvariant_with_p2v_extension})
% worked out Section \ref{subsec:identitymappings}.

%\begin{figure}\footnotesize
%  \begin{lstlisting}[mathescape,escapeinside={(*}{*)}]
%  /*... Continued from Figure (*\ref{fig:calltopteinitialize}*); assertions below specialized to non-allocating path for clarity */
%  $\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus \{ \mathsf{entry} \}),m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{rcx}  \mapsto_{\textsf{r}} \textsf{\_}  \ast \textsf{entry} \mapsto_{\textsf{id}} \textsf{\_} \ast \ghostmaptoken{\delta{}s}{\rtv}{\delta}  }_{\rtv}$
%  $\specline{ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{(pte\_initialized (entry\_val.pfn))} \urcorner }_{\rtv}$
%  $\specline{ \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{(pte\_initialized (entry\_val.pfn)))} \ast \texttt{rax} \mapsto_{\textsf{r}} \textsf{ table\_root (pte\_initialize(entry\_val.pfn))} }_{\rtv}$
%  $\specline{\forall_{i\in \textsf{0 ... 511} } \ldotp  \ghostmaptoken{\textsf{id}}{((\textsf{table\_root (pte\_initialized (entry\_val.pfn)))}) + \textsf{i * 8})}{\textsf{v-1}}  }$ (*\label{line:children}*)
%  uintptr_t next_phys_addr = PTE_PFN_TO_ADDR(entry->pfn)
%  uintptr_t next_virt_addr = (uintptr_t) P2V(next_phys_addr); (*\label{line:p2v}*) 
%  /*... clean up the stack and return the rax value*/
%  return next
%}
%\end{lstlisting}
%\vspace{-1em}
%\caption{Converting a physical address of a PTE to a virtual address (w/o instruction pointer or flag updates).
% % Abreviating the other relevant resources (including the lower entries) peeled-out of the invariant as $\textsf{R}_{\textsf{children}}$
%\vspace{-1em}
%}
%\label{fig:p2v}
%\end{figure}
Lines \ref{line:childrenC}-\ref{line:p2vC} in Figure \ref{fig:calltopteinitializeC} show the verification of the end of \lstinline|pte_get_next_table| specialized to the case where 
where no allocation was necessary (i.e., the conditional on Line \ref{line:check_entry_presentC} of Figure \ref{fig:calltopteinitializeC} was taken).
In this case, the true present bit allows access to the child tokens from Line \ref{line:conditional_childrenC} of Figure \ref{fig:calltopteinitializeC},
which is then refined to the assertion on Line \ref{line:childrenC} of Figure \ref{fig:calltopteinitializeC}.
The code loads the value at the offseted (\textsf{KERNBASE}) virtual memory address ($\textsf{entry}_{\textsf{pfn}}$ \textsf{+KERNBASE})
of the \emph{base} of the next level of the page table.
% \todo[inline]{the next sentence depends on having figure 10 updated to reflect the page-worth of tokens}
While we could now convert this address to a virtual points-to, this is not necessarily the correct thing to do.
The caller \lstinline|walkpgdir| (discussed next) uses \lstinline|pte_get_next_table| to retrieve just the base address,
because only the caller knows which entry in the subsequent table will be accessed (it depends on the corresponding bits from the virtual
address being translated). So instead we pass back the per-address-space invariant with the identity mapping resources for \lstinline|entry|
pulled out. The caller determines which entry in that table must actually
be accessed --- by selecting the appropriate index into the 512 ghost map tokens returned in the postcondition,
and using the ghost translation and physical location portions of the invariant to assemble a vpte-pointsto
that justifies the caller's subsequent access to a particular entry of the returned table.
% in the identity map ($\Xi\setminus\{entry\}$) of the kernel invariant.
% the logical update in Specification  Lines 5-10 to 10-14 for obtaining virtual-pointsto resource for the frame 
% ($\textsf{entry}_{\textsf{pfn}}$) by removing it from the ghost map ($\Xi\setminus\{entry\}\cup \{\textsf{entry}_{\textsf{pfn}}) \}$) 
% in Line 5 and compute the identity mapping for this physical frame address in Line 13 in Figure \ref{fig:p2v}).
